Classless Java: Tuning Java Interfaces

### 2015-12-02 paper/sections/CaseStudy.tex
Case studies

\section{Case Studies}
In this section we conduct three case studies which reveals various
advantages using \mixin annotation. The first case study provides a
simple way to solve the Expression Problem, while supporting multiple,
independent extensions in Java.  The second case study show how to model an
embedded DSL for SQL languages with fluent interfaces. Finally, the third 
case study models a simple game, and compares our implementation with an 
existing one, showing that the amount of code is reduced significantly using \mixin.

@@

The \emph{Expression Problem} (EP)~\cite{wadler98expression} is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement is
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility: the ability to assemble a system
from multiple, independently developed extensions. Unfortunatelly, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing.

\paragraph{Initial System}
In the formulation of the EP there is an initial system that models
arithmetic expressions with only literals and addition, and an initial 
operation \texttt{eval} for expression evaluation.

\noindent As shown in Figure~\ref{fig:ep}, \texttt{Exp} is the common
super-interface with an evaluation operation \texttt{eval()}
inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend interface
\texttt{Exp} with default implementations for the \texttt{eval} operation. The
number field \texttt{x} of a literal is represented as a getter method
\texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition
as getter methods \texttt{e1()} and \texttt{e2()}.

\begin{figure}
\begin{tabular}{l|l}
\lstinputlisting[linerange=34-41]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_INIT
&
\lstinputlisting[linerange=50-56]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
\end{tabular}
\caption{The Expression Problem. Initial system on the left, and code for adding
  print operation on the right. }\label{fig:ep}
\end{figure}

@@

\paragraph{Independent Extensibility}
To show that our approach supports independent extensibility~\cite{zenger05independentlyextensible}, we first define a
new operation \texttt{collectLit} (which collects all
literal components in an expression) on expressions. For space reasons, 
we omit the definitions of the methods:

\begin{lstlisting}[]
interface ExpC extends Exp { List<Integer> collectLit(); }
@Obj interface LitC extends Lit, ExpC {
    default List<Integer> collectLit() { ... }}
@Obj interface AddC extends Add, ExpC {
    ExpC e1(); ExpC e2(); //return type refined!
    default List<Integer> collectLit() { ... }}
\end{lstlisting}

@@

Note that the code for instantiation is automatically generated by \mixin. 
So, for example creating a simple expression of type \texttt{ExpPC} is 
as simple as:
\begin{lstlisting}
ExpPC e8 = AddPC.of(LitPC.of(3), LitPC.of(4));
\end{lstlisting}
\noindent In contrast, in a pure Java solution, the tedious instantiation code would need 
to be defined manually.

@@

The \mixin annotation can also be used to create fluent interfaces.  When
creating fluent interfaces with \mixin, there are two main advantages:
\begin{enumerate}
\item Instead of forcing programmers to hand write code using \textbf{return
    this}, our approach with \mixin annotation removes this verbosity and
  automatically generate fluent setters.
\item Along the extension direction, the return types of fluent setters are
  automatically refined.
\end{enumerate}

@@

\noindent The \mixin annotation allowed us to reduce the interfaces/classes used
in Bono et al.'s implementation by 21.4\% (from 14 to 11). The
reductions was due to the replacement of instantiation classes with
generated \texttt{of} methods. The number of source lines of code (SLOC)
was reduced by 40\% due to both the removal of instantiation overhead and
generation of getters/setters. %\bruno{explain why? we generate code for getters/setters?}. 
To ensure a fair comparison, we used the same coding style as Bono et al.'s.

>>>>

\section{Case Studies}
In this section we conduct three case studies which reveal various
advantages of using the \mixin annotation. The first case study provides a
simple way to solve the Expression Problem, while supporting multiple,
independent extensions in Java.  The second case study show how to model an
embedded DSL for SQL languages with fluent interfaces. Finally, the third 
case study models a simple game, and compares our implementation with an 
existing one, showing that the amount of code is reduced significantly using \mixin.

@@

The \emph{Expression Problem} (EP)~\cite{wadler98expression} is a well-known
problem about modular extensibility issues in software evolution. Recently, a
new solution~\cite{eptrivially} using only covariant type refinement was
proposed. When this solution is modeled with interfaces and default methods, it
can even provide independent extensibility~\cite{zenger05independentlyextensible}: the ability to assemble a system
from multiple, independently developed extensions. Unfortunatelly, the
required instantiation code makes a plain Java solution verbose and cumbersome
to use. The \mixin annotation is enough to remove the boilerplate code, making
the presented approach very appealing.

\paragraph{Initial System}
In the formulation of the EP there is an initial system that models
arithmetic expressions with only literals and addition, and an initial 
operation \texttt{eval} for expression evaluation.
As shown in Figure~\ref{fig:ep}, \texttt{Exp} is the common
super-interface with an evaluation operation \texttt{eval()}
inside. Sub-interfaces \texttt{Lit} and \texttt{Add} extend interface
\texttt{Exp} with default implementations for the \texttt{eval} operation. The
number field \texttt{x} of a literal is represented as a getter method
\texttt{x()} and expression fields (\texttt{e1} and \texttt{e2}) of an addition
as getter methods \texttt{e1()} and \texttt{e2()}.

\begin{figure}
\begin{tabular}{l|l}
\lstinputlisting[linerange=34-41]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_INIT
&
\lstinputlisting[linerange=50-56]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_PRINT
\end{tabular}
\caption{The Expression Problem. Initial system on the left, and code for adding
  print operation on the right. }\label{fig:ep}
\end{figure}\bruno{Yanlin: code is overflowing the margins. Note that
  there are 2 spare lines that can be used on the right!}

@@

\paragraph{Independent Extensibility}
To show that our approach supports independent extensibility, we first define a
new operation \texttt{collectLit}, which collects all
literal components in an expression. For space reasons, 
we omit the definitions of the methods:

\begin{lstlisting}[]
interface ExpC extends Exp { List<Integer> collectLit(); }
@Obj interface LitC extends Lit, ExpC {...}
@Obj interface AddC extends Add, ExpC {ExpC e1(); ExpC e2(); ...}
\end{lstlisting}

@@

Note that the code for instantiation is automatically generated by \mixin. 
Creating a simple expression of type \texttt{ExpPC} is 
as simple as:
\begin{lstlisting}
ExpPC e8 = AddPC.of(LitPC.of(3), LitPC.of(4));
\end{lstlisting}
\noindent In contrast, in a pure Java solution, the tedious instantiation code would need 
to be defined manually.

@@

The \mixin annotation can also be used to create fluent interfaces.  When
creating fluent interfaces with \mixin, there are two main advantages:
\begin{enumerate}
\item Instead of forcing programmers to hand write code using \textbf{return
    this}, our approach with \mixin annotation removes this verbosity and
  automatically generate fluent setters.
\item The approach supports extensibility: the return types of fluent setters are
  automatically refined.
\end{enumerate}

@@

\noindent The \mixin annotation allowed us to reduce the interfaces/classes used
in Bono et al.'s implementation by 21.4\% (from 14 to 11). The
reductions were due to the replacement of instantiation classes with
generated \texttt{of} methods. The number of source lines of code (SLOC)
was reduced by 40\% due to both the removal of instantiation overhead and
generation of getters/setters. %\bruno{explain why? we generate code for getters/setters?}. 
To ensure a fair comparison, we used the same coding style as Bono et al.'s.
